Visual narratives of the Covid-19 pandemic

### 2022-01-17 JDSSV_manuscript_knitr.tex
more thoughts

\documentclass[article]{jdssv}

@@

\usepackage{subcaption}

@@

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
    email: \email{susan.vanderplas@unl.edu}
  Susan Vanderplas\\
  Department of Statistics\\
  University of Nebraska–Lincoln\\
  349A Hardin Hall \\
  Lincoln, NE 68583-0963, USA\\
  E-mail: \email{susan.vanderplas@unl.edu}\\
  URL: \url{https://statistics.unl.edu/susan-vanderplas}\\
\ \\
%}
%
%\Address{
  Adalbert F.X. Wilhelm\\
  Department of Psychology and Methods\\
  Jacobs University Bremen gGmbH\\
  Campus Ring 1\\
  28759 Bremen, Germany\\
  E-mail: \email{a.wilhelm@jacobs-university.de}\\
  URL: \url{http://www.jacobs-university.de/directory/wilhelm}
}

\begin{document}
% \SweaveOpts{concordance=TRUE}
<<include=FALSE, cache=FALSE, echo=FALSE, purl=FALSE>>=
library(knitr)
opts_chunk$set(comment = NA, size = 'footnotesize', prompt = TRUE, highlight = FALSE, 
               cache = TRUE, crop = TRUE, concordance = FALSE, fig.align='center',
               fig.path='Figures_R/fig-',out.width="0.4\\textwidth", 
               message = F, error = F, warning = F, echo = F)
opts_knit$set(out.format = "latex")
knit_hooks$set(crop = hook_pdfcrop)
options(width = 80, prompt = "R> ", continue = "    ", digits = 3)
thm <- knit_theme$get('default')
Sys.setenv(TEXINPUTS=getwd(),
           BIBINPUTS=getwd(),
           BSTINPUTS=getwd())
knit_theme$set(thm)
@

<<get-nyt-data, include = F>>=

# Read in the data from NYT
library(tidyverse)
library(lubridate)
library(zoo)
library(ggrepel)
library(gridExtra)
cvstate <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")

cvstate <- cvstate %>%
  group_by(state, fips) %>%
  mutate(date_100th_case = min(date[cases >= 100]),
         days_100th_case = date - date_100th_case,
         daily_cases = cases - lag(cases, 1, default = 0),
         week_avg_cases = rollmean(cases, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
         daily_avg_cases = rollmean(daily_cases, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
         daily_deaths = deaths - lag(deaths, 1, default = 0),
         week_avg_deaths = rollmean(deaths, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T),
         daily_avg_deaths = rollmean(daily_deaths, 7, align = "right", fill = c(0, NA, NA), na.pad = T, na.rm = T)) %>%
  ungroup() %>%
  mutate(state = factor(state))

download.file("https://www2.census.gov/programs-surveys/popest/tables/2010-2019/state/totals/nst-est2019-01.xlsx", destfile = "census_pop_state.xlsx", quiet = T, mode = "wb")
pop_data <- readxl::read_xlsx("census_pop_state.xlsx", col_names = c("state", "pop2019"), col_types = c("text", rep("skip", 11), "numeric"), skip = 9) %>%
  mutate(state = str_remove(state, "[[:punct:]]"))

cvstate <- left_join(cvstate, pop_data, by = "state")

my_theme <- theme_bw() + 
  theme(plot.background = element_rect(fill = "transparent", colour = NA), 
        panel.background = element_rect(fill = "transparent", colour = NA), 
        legend.background = element_rect(fill = "transparent", colour = NA),
        axis.text.y = element_text(angle = 90, hjust = 0.5)) 
theme_set(my_theme)
@

<<colourpal, include = F>>=
# Including this for temporary compatibility - we may want to use *shudder* viridis?
main_colours <- c("#d00000", "#f5f1e7", "#c7c8ca")

# use as accents
secondary_colours <- c("#001226", "#249ab5")

# These are only to be used for infographics/charts
tertiary_colours <- c("#bccb2a", "#f58a1f", "#005d84", "#ffd74f", "#a5228d")

colour_palette <- c("#d00000", "#249ab5",  "#1B8500", "#f58a1f", "#a5228d", "#001226", "#B1EB66", "#005d84", "#c25700")

heat_colour_palette <- c("#ffd74f", "#f58a1f", "#c25700", "#d00000", "#9D0000", "#6A0000")

blueseqpal <- c("001226", "06293E", "0C3F56", "12566E", "186D85", "1E839D", 
                "249AB5", 
                "39ABC1", "4EBCCE", "64CDDA", "79DDE6", "8EEEF3", "A3FFFF") %>% 
  paste0("#", .) %>% rev()

blueseqpal_lt <- rev(c("#B2FFFF", "#C2FFFF", "#D1FFFF", "#E0FFFF", "#F0FFFF"))

darkblueseqpal <- colorspace::darken(blueseqpal, .2)
lightblueseqpal <- colorspace::lighten(blueseqpal, .2)
@
%% -- Introduction -------------------------------------------------------------

@@

\begin{figure}
\centering
\begin{subfigure}[t]{.45\textwidth}
\includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state, per million people. This somewhat mitigates the area bias issue, but is still subject to other perceptual biases. Source: The COVID Tracking Project at The Atlantic \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
\end{subfigure}\hfill
\begin{subfigure}[t]{.45\textwidth}
\includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
\caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
\end{subfigure}
\end{figure}

Throughout the pandemic, both in choropleth maps and time-series graphs, creators have had to grapple with scale. When scales automatically adjust, mental comparisons across time can become misleading, because e.g. the colour scale from yesterday does not mean the same thing as it did today, as the range has expanded and the colours have stayed the same. This problem is more noticeable in choropleths than in time-series charts, where we are more likely to expect and notice a scale change, but it is still a fundamental issue in many different types of charts. Some publications leaned into this by arranging their graphics to highlight the fact that cases were ``off the charts", as in \Cref{fig:jbm-tweet}, implicitly cuing readers that the situation was exceptional and worthy of their attention.

\begin{figure}
\includegraphics[width=.99\linewidth]{ft-off-charts}
\caption{Chart showing the increase in cases as a result of the B.1.617.2 variant in regions of the UK. Notably, this chart arranges areas to accommodate certain regions whose cases are ``off the chart" in order to emphasize the exceptional situation. Source: Tweet by John Burn-Murdoch (@jburnmurdoch), designer for the Financial Times. \url{https://twitter.com/jburnmurdoch/status/1397995388267810818}.}
\label{fig:jbm-tweet}
\end{figure}

% Is it worth discussing e.g. https://www.nytimes.com/interactive/2021/us/covid-risk-map.html that does not make this distinction and runs into the "boy who cried wolf" problem? When everything is always extremely high risk we just tend to tune things out. I don't want to distract from the overview, though... https://twitter.com/rypan/status/1282020934044401664 is a good example

@@

\section{Time series}
% Line plots, frequency domain, NY times tile plots w/ cases represented in colour intensity

Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis. Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis. This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. Time series plots are the most natural choice for these goals, and over the course of the pandemic, many different attempts have been made to show time series information using different graphical forms.

\begin{figure}
\centering
\includegraphics[width=.99\linewidth]{wapo-covid-monitoring}
\caption{COVID monitoring summary charts on the Washington Post's Coronavirus page. Source: \url{https://www.washingtonpost.com/coronavirus/}, 2022-01-17.}
\label{fig:wapo-covid-time-series}
\end{figure}

The most obvious time-series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-covid-time-series}. This allows for individuals to assess the current situation and easily compare the current number of cases to past times for which the individual has a direct reference for what to expect in impact to daily life. Often, as in \Cref{fig:wapo-covid-time-series}, these time series show multiple variables: cases, deaths, and hospitalizations (and sometimes vaccinations) to provide a more comprehensive view of the situation; this is particularly useful as there has been some decoupling of case numbers and hospitalizations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe. Many outlets also provide additional charts designed to provide some immediate context to readers, as shown in \Cref{fig:wapo-context}.

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{wapo-covid-context}
\caption{Some outlets provide contextual information to assist readers with interpreting time-series plots. Source: Washington Post Coronavirus Tracking, \url{https://www.washingtonpost.com/graphics/2020/national/coronavirus-us-cases-deaths/}.}
\label{fig:wapo-context}
\end{figure}

However, there are some limitations with this presentation: if we are interested in showing the change in counts over time, viewers must assess the slope of the line, rather than its position. We know that slope judgments are much less accurate than position \citep{clevelandGraphicalPerceptionVisual1987}, even in relatively simple situations; when we add in our ability to assess exponential growth, our perceptual accuracy is even more suspect.

Other time-series representations were more abstract, sacrificing exact data representations for a higher-level summary overview that allowed viewers to make comparisons between regions and points in time without the burden of numerical calculations. Several versions from the New York Times are shown in \Cref{fig:sparklines-heatmap-nyt}. The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data, however, they do not present precise numerical information and even the color scale is fairly opaque -- it would be extremely difficult to translate the visualization into any sort of accurate numerical estimate. In addition, by hiding these details from the user, it is very difficult to identify when the representation changes over time, as it did several times during the summer of 2020. The use of a similar color scheme as well as similar representations made it extremely difficult for users to identify that the value represented had changed. Clearly, the purpose of these charts is not to provide numerical precision, but rather a comparative assessment of the status of one region relative to others as well as relative to previous time points.

\begin{figure}
\centering
\includegraphics[width=.3\linewidth]{NYT-heatmap-crop}\hfill
\includegraphics[width=.3\linewidth]{NYT-heatmap-crop3}\hfill
\includegraphics[width=.3\linewidth]{NYT-heatmap-crop2}
\caption{Sparklines style heatmaps embedded in a table of coronavirus data by state. Over the course of Summer 2020, the specific measures shown changed, even though the form of the data remained highly similar. Source: \url{https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html} (May 6, 2020; June 15, 2020; and July 24, 2020, respectively) }

\label{fig:sparklines-heatmap-nyt}
\end{figure}

One unique time-series plot created a representation of the case counts in the frequency domain: instead of plotting cases as they occurred, the chart instead shows line segments from 0 to N in $y$, where $N$ is a predefined number of cases; then the chart resets to show the next angle. This produces a sense of intensity of the waves of the pandemic over time as shown in \Cref{fig:freq-domain}. Interestingly, this frequency representation can also be easily transferred into an audio domain, providing access for visually impaired users as well as a multimodal representation of the data for sighted users.

\begin{figure}
\centering
\includegraphics[width=.75\linewidth]{frequency-domain}
\caption{Frequency-domain case count graphics. Here, each 1000 deaths are represented by a single line segment whose slope represents the time taken to reach the next 1000 deaths. This provides viewers with a sense of the pace of the epidemic, rather than the raw case counts in more standard time-series representations. This representation can also be shown in the auditory domain, providing access to those who are typically excluded from visualizations due to vision loss. Source: Romain Vuillemot, \url{https://observablehq.com/@romsson/sonification-of-covid-19-data}.}
\label{fig:freq-domain}
\end{figure}

There are many different ways that designers can use time-series data to provide additional contextual information and facilitate comparisons. \Cref{fig:ft-waves} shows covid cases, positivity rate, and hospital admissions from South Africa's Gauteng province, which was one of the first areas to experience a large wave of the Omicron variant. This chart provides context for the Omicron wave, showing that while infections are occurring at a much faster rate than in previous waves, hospital admissions seem to be approximately following previous trends.

\begin{figure}\centering
\includegraphics[width=\linewidth]{ft-waves}\\
\includegraphics[width=\linewidth]{ft-waves-linear}
\caption{Covid Cases, positivity rate, and hospital admissions from South Africa's Gauteng province, on a log scale (top) and linear scale (bottom). Successive waves are overlaid, showing that the increase in cases due to the Omicron variant was sharper than any previously measured increase, while hospitalizations were much more comparable to previous waves. Source: John Burn-Murdoch, \url{https://twitter.com/jburnmurdoch/status/1466480113487392769}}
\label{fig:ft-waves}
\end{figure}

While traditionally, time-series information has been limited to charts with time on the x-axis, there have been many attempts to make use of the capabilities of web graphics and animation in order to convey temporal information in less traditional time-series forms. One controversial chart appeared in the New York Times; the x-axis was wrapped around an Archimedean spiral to provide a sense of periodicity in the year-by-year evolution of case counts, as shown in \Cref{fig:nyt-spiral}. The original form of this chart is prone to the line-width illusion \citep{vanderplasSignsSineIllusion2015}, in addition to the well-known problems with polar charts \citep{hofmannGraphicalTestsPower2012,waldnerComparisonRadialLinear2020}. The re-envisioned version, which aligns counts on the outside of the spiral, and displays them as radial lines, mitigates some of the issues with the line-width illusion by directly showing each line as its own entity, facilitating direct comparisons. In addition, the re-envisioned chart includes a reference line at 100K cases/day, which allows the reader to compare the severity of different waves directly. Still, it is more work to interpret and compare peaks on this chart than on a similar linear time series chart with the same data, as the reader must assess line width and then do a mental rotation and shift operation in order to compare with any other time period.

\begin{figure}\centering
\includegraphics[width=.45\linewidth]{nyt-spiral}\hfill\includegraphics[width=.45\linewidth]{spiral-rework}
\caption{Covid cases in the United States, 2020-2022. The original graph from the New York Times is on the left, and a reenvisioned version which is more perceptually friendly is on the right. Source: (Left) New York Times \url{https://www.nytimes.com/2022/01/06/opinion/omicron-covid-us.html}, (Right) Covid Spiral, by Sourya Shrestha, \url{https://github.com/souryashrestha/covid_spiral}.}
\label{fig:nyt-spiral}
\end{figure}

% bar-chart race https://www.youtube.com/watch?v=8WH-bTH6GqI

% BBC flowers w/ sonification https://www.bbc.co.uk/news/resources/idt-7464500a-6368-4029-aa41-ab94e0ee09fb

% Other animations?

@@

\section{To log or not to log}
As COVID cases grow quasi-exponentially while there are susceptible members of the population (subject to the effectiveness of mitigation measures and testing availability), it seems natural to use log scales to allow for more effective comparisons of slight changes in case counts over time. In addition, log scales make it possible to compare regions with different populations or infection rates in the same chart. As noted previously, however, interpreting log scales requires levels of numerical sophistication that may not be appropriate for the general public. Even researchers do not always read and interpret log scales correctly\citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so is difficult under normal circumstances\citep{hecklerStudentAccuracyReading2013} is difficult. When panic, fear, uncertainty, and doubt about the situation are added to the mix, it is easy to imagine that we become even worse when interpreting graphics.
% Discussion of how this section's conclusions change with the pandemic

@@

<<log-scale-initial, fig.width = 8, fig.height = 4, out.width = "\\linewidth", fig.cap = "In the early stages of the pandemic, log scales allowed the comparison of raw case counts in locations with vastly different population and case counts.">>=
# remotes::install_github("AllanCameron/geomtextpath")
library(geomtextpath)

library(facetscales)
scales_y <- list(
  Log = scale_y_log10(),
  Linear = scale_y_continuous()
)
tmp <- cvstate %>% 
  filter(date < ymd("2020-06-01")) %>%
  filter(state %in% c("Michigan", "Washington", "New York", "California"))

p1 <-  tmp %>%
  ggplot(aes(x = date, y = daily_avg_cases, group = state)) +
  geom_textline(aes(colour = state, label = state, vjust = -0.1, hjust =.93) , text_smoothing = 50) + 
  ylab("Daily average cases (7 day average)") + 
  xlab("Date") + 
  guides(colour = "none") + 
  scale_y_log10() + 
  facet_grid(~"Log") + 
  theme_bw()

p2 <-  tmp %>%
  ggplot(aes(x = date, y = daily_avg_cases, group = state)) +
  geom_textline(aes(colour = state, label = state, vjust = -0.25, hjust = .85) , text_smoothing = 60) + 
  ylab("Daily average cases (7 day average)") + 
  xlab("Date") + 
  guides(colour = "none") + 
  facet_grid(~"Linear") + 
  theme_bw() +
  theme(axis.title.y = element_blank())

grid.arrange(p1, p2, nrow = 1)
@

@@

% Problems with log scales
However, after the first wave of COVID, the issues with log scales became more apparent: it was difficult to detect slight increases in case counts that indicated the beginning of a new wave amid a background level of spread, as demonstrated in  \Cref{fig:log-scale-failures}. Diagonal reference lines from the origin were also less helpful, as the growth of cases or deaths was no longer approximately exponential and varied over time; for these reference lines to be effective there would need to be a clear idea of when the case counts started to increase exponentially, which is difficult to determine whilst in the thick of a potential COVID wave. Occasionally graphs with manually drawn reference lines were made available, but usually only in a retrospective manner, as in \Cref{fig:log-scale-ref-multiple}.

<<log-scale-failures, echo = F, fig.width = 8, fig.height = 4, out.width = ".95\\linewidth", message = F, error = F, warning = F, fig.cap = "One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread. Linear scales do not have this problem - the exponential source is noticeable very quickly in the total line, but on the log scale it is much harder to discern when the exponential source causes the total line to diverge from the background. In the top-right corner, it is difficult to identify that there is an exponential increase in cases amid the baseline, even though the exponential source makes up approximately 50\\% of the cases at the end of the time period shown." >>=
# Set up labeller to tweak things 
rlistmap <- function(x, myfun) {
  if ("list" %in% class(x)) return(purrr::map(x, ~rlistmap(., myfun)))

  return(myfun(x))
}
fix_labels <- function(x) {
  x  %>%
    str_replace('scale: (.*)', "\\1 scale") %>%
    str_replace("baseline: (\\d{1,})", "Baseline:\n\\1 cases")
}
label_custom <- function(labels, multi_line = T, sep = ": ") {
  out <- label_both(labels, multi_line = multi_line, sep = sep)

  out <- rlistmap(out, fix_labels)

  out
}
class(label_custom) <- "labeller"

# Set up data 
growth <- rpois(30, 15*exp((1:30 - 20)/5))

tmp <- tibble(baseline = 15,
       x = 1:30,
       y = rpois(30, 15),
       y1 = growth) %>%
  bind_rows(
    tibble(baseline = 50,
       x = 1:30,
       y = rpois(30, 50),
       y1 = growth) 
  ) %>%
  bind_rows(
    tibble(baseline = 200,
       x = 1:30,
       y = rpois(30, 200),
       y1 = growth) 
  ) %>%
  pivot_longer(matches("y")) %>%
  mutate(name = str_replace_all(name, c("y$" = "Background", "y1" = "Exponential")))  %>%
  group_by(baseline, x) %>%
  mutate(Total = sum(value)) %>%
  pivot_wider(names_from = "name", values_from = "value") %>%
  pivot_longer(matches("Background|Exponential|Total")) %>%
  mutate(name = factor(name, levels = c("Total", "Exponential", "Background")))

tmp <- bind_rows(mutate(tmp, scale = "Log"), mutate(tmp, scale = "Linear")) %>%
  mutate(Baseline = reorder(paste0("Baseline: ", baseline, " cases"), baseline)) %>%
  mutate(scale = factor(scale, levels = c("Log", "Linear")))

# Set up scales 
scales_y <- list(
  Log = scale_y_log10(),
  Linear = scale_y_continuous()
)

tmp %>%
ggplot(aes(x = x, y = value, colour = name, size = name)) + 
  geom_line(alpha = .9) +
  facet_grid_sc(rows = vars(scale), cols = vars(Baseline), 
                scales = list(x = "fixed", y = scales_y)) +
  scale_size_manual(values = c(1.75, 1, 1)) +
  scale_colour_manual(values = c("#249ab5", "#d00000", "#001226")) +
  scale_fill_discrete(name = "") + 
  theme_bw() + 
  theme(legend.position = "bottom", axis.title = element_blank(),
        legend.title = element_blank())
@

% Focus primarily on time-series graphs (as opposed to log scales in colour)

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{ft-reference-lines}
\caption{An annotated time series from the Financial Times which appeared on Twitter on November 12, 2020. This chart has annotations which show the decreasing $R_{eff}$ in successive peaks of the pandemic, with periods of $R_{eff}<1$ between surges in cases. These annotations assist the reader with drawing complex conclusions from the chart, but are difficult to automate and thus tend to be manually curated. Source: Mark Gubrud (@mgubrud), \url{https://twitter.com/mgubrud/status/1326784082399944704}.}
\label{fig:log-scale-ref-multiple}
\end{figure}

@@

<<linear-scales-ref-lines, fig.width = 8, fig.height = 4, out.width = ".8\\linewidth", fig.cap = "Reported COVID cases in New York State, 2020-2022. The linear scale makes it difficult to compare the trajectory of different waves to determine how severe the current status is relative to the past, because the primary contrast is the height of the relative peaks, rather than the growth \\emph{rate}. A similar graph on the log scale would have the peaks at much more similar heights (though there would still be a difference), allowing the reader to focus on other information, such as the slope of the relative lines.">>=
cvstate %>%
  filter(state == "New York") %>%
  ggplot(aes(x = date, y = daily_avg_cases)) + 
  geom_line() + 
  ggtitle("Reported COVID cases in New York State, 2020-2022") +
  ylab("Daily Cases reported (7-day average)") + 
  xlab("Date") + theme_bw()
@

@@

It is not clear that the use of log or linear scales during the COVID-19 pandemic had a large effect on public opinion. Several studies were conducted in the early stages of the pandemic \citep{romanoScaleCOVID19Graphs2020, seviLogarithmicLinearVisualizations2020, ryanGraphsLogarithmicAxes2020} and results seem to suggest that while individuals have difficulty understanding log scale graphs, these issues do not tend to affect their support for intervention measures, perhaps in part because COVID-related news saturated the news and opinions were set outside of information provided in the experiments. This saturation makes it difficult to study the graphical influences while removing effects of popular opinion, political leaning, and emotional sentiment relating to the pandemic.

If we evaluate the use of log and linear scales under the more general context of exponential (or near-exponential) growth rates, however, we can gain some clarity as to their use in this particular context. We are abysmal at forecasting exponential growth, vastly under-estimating future growth by using linear or quadratic approximations \citep{wagenaarExtrapolationExponentialTime1978a,lawrenceExploringJudgementalForecasting1992,timmersInverseStatisticsMisperception1977a}. If the goal of a chart of COVID case counts is to allow individuals to forecast the trend and make decisions accordingly, then using a log scale (with all of its pitfalls in understanding) may be the best option, as it at least replaces the need to forecast along an exponential curve with the need to forecast along a straight line. However, it is not clear that individuals can transfer that prediction back to a linear scale. If the goal of presenting a chart of case counts is to tell the story of what has happened (rather than supporting future decision-making), then it is undoubtedly better to present the data on a more familiar linear scale that requires less cognitive load on the part of the viewer.

Unfortunately, there are not a lot of good options for supporting forecasting decisions in mathematically unsophisticated viewers. Guide lines, like those used in the Financial Times, were helpful in the initial phases of the pandemic, but no outlets that we are aware of shifted the lines to show the growth rate of the current peak (as opposed to the initial peak), and due to the issues shown in \Cref{fig:log-scale-failures}, these guide lines may not have been all that successful in any case. Those hoping to create successful time-series charts that allow forecasting are thus left with two options: deal with our inability to forecast exponential growth, or deal with most individuals' inability to translate log scales into practical reality.

@@

If necessary or useful, information about certain computational details
such as version numbers, operating systems, or compilers could be included
in an unnumbered section. Also, auxiliary packages (say, for visualisations,
maps, tables, \dots) that are not cited in the main text can be credited here.

>>>>

\documentclass[article]{jdssv}\usepackage[]{graphicx}\usepackage[]{color}
% maxwidth is the original width if it is less than linewidth
% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}

@@

@@

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
    email: \email{susan.vanderplas@unl.edu}
  Susan Vanderplas\\
  Department of Statistics\\
  University of Nebraska–Lincoln\\
  349A Hardin Hall \\
  Lincoln, NE 68583-0963, USA\\
  E-mail: \email{susan.vanderplas@unl.edu}\\
  URL: \url{https://statistics.unl.edu/susan-vanderplas}\\
\ \\
%}
%
%\Address{
  Adalbert F.X. Wilhelm\\
  Department of Psychology and Methods\\
  Jacobs University Bremen gGmbH\\
  Campus Ring 1\\
  28759 Bremen, Germany\\
  E-mail: \email{a.wilhelm@jacobs-university.de}\\
  URL: \url{http://www.jacobs-university.de/directory/wilhelm}
}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
% \SweaveOpts{concordance=TRUE}

%% -- Introduction -------------------------------------------------------------

@@

\section{Time series and dynamics of the outbreak}
% Line plots, frequency domain, NY times tile plots w/ cases represented in color intensity

@@

@@

\section{To log or not to log}
As COVID cases grow quasi-exponentially while there are susceptible members of the population (subject to the effectiveness of mitigation measures and testing availability), it seems natural to use log scales to allow for more effective comparisons of slight changes in case counts over time. In addition, log scales make it possible to compare regions with different populations or infection rates in the same chart. As noted previously, however, interpreting log scales requires levels of numerical sophistication that may not be appropriate for the general public. Even researchers do not always read and interpret log scales correctly\citep{mengeLogarithmicScalesEcological2018}; expecting the general public to do so is difficult under normal circumstances\citep{hecklerStudentAccuracyReading2013} is difficult. When panic, fear, uncertainty, and doubt about the situation are added to the mix, 
% Discussion of how this section's conclusions change with the pandemic

@@

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{0.961, 0.961, 0.961}\color{fgcolor}\begin{figure}

{\centering \includegraphics[width=\linewidth]{Figures_R/fig-log-scale-initial-1}

}

\caption[In the early stages of the pandemic, log scales allowed the comparison of raw case counts in locations with vastly different population and case counts]{In the early stages of the pandemic, log scales allowed the comparison of raw case counts in locations with vastly different population and case counts.}\label{fig:log-scale-initial}
\end{figure}

\end{knitrout}

@@

% Problems with log scales
However, after the first wave of COVID, the issues with log scales became more apparent: it was difficult to detect slight increases in case counts that indicated the beginning of a new wave amid a background level of spread, as demonstrated in  \Cref{fig:log-scale-failures}. Diagonal reference lines from the origin were also less helpful, as the growth of cases or deaths was no longer approximately exponential and varied over time; for these reference lines to be effective there would need to be a clear idea of when the case counts started to increase exponentially, which is difficult to determine whilst in the thick of a potential COVID wave.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{0.961, 0.961, 0.961}\color{fgcolor}\begin{figure}

{\centering \includegraphics[width=.95\linewidth]{Figures_R/fig-log-scale-failures-1}

}

\caption[One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread]{One problem with log scales is that if there is a background level of spread, it can be hard to notice the introduction of an additional source of exponential spread. Linear scales do not have this problem - the exponential source is noticeable very quickly in the total line, but on the log scale it is much harder to discern when the exponential source causes the total line to diverge from the background. In the top-right corner, it is difficult to identify that there is an exponential increase in cases amid the baseline, even though the exponential source makes up approximately 50\% of the cases at the end of the time period shown.}\label{fig:log-scale-failures}
\end{figure}

\end{knitrout}

% Focus primarily on time-series graphs (as opposed to log scales in color)

@@

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{0.961, 0.961, 0.961}\color{fgcolor}\begin{figure}

{\centering \includegraphics[width=.8\linewidth]{Figures_R/fig-linear-scales-ref-lines-1}

}

\caption[Reported COVID cases in New York State, 2020-2022]{Reported COVID cases in New York State, 2020-2022. The linear scale makes it difficult to compare the trajectory of different waves to determine how severe the current status is relative to the past, because the primary contrast is the height of the relative peaks, rather than the growth \emph{rate}. A similar graph on the log scale would have the peaks at much more similar heights (though there would still be a difference), allowing the reader to focus on other information.}\label{fig:linear-scales-ref-lines}
\end{figure}

\end{knitrout}

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{0.961, 0.961, 0.961}\color{fgcolor}\begin{figure}

{\centering \includegraphics[width=.8\linewidth]{Figures_R/fig-log-scales-ref-lines-1}

}

\caption[Reported COVID cases in New York State, 2020-2022]{Reported COVID cases in New York State, 2020-2022. The linear scale makes it difficult to compare the trajectory of different waves to determine how severe the current status is relative to the past.}\label{fig:log-scales-ref-lines}
\end{figure}

\end{knitrout}

@@

It is not clear that the use of log or linear scales during the COVID-19 pandemic had a large effect on public opinion. Several studies were conducted in the early stages of the pandemic \citep{romanoScaleCOVID19Graphs2020, seviLogarithmicLinearVisualizations2020, ryanGraphsLogarithmicAxes2020} and results seem to suggest that while individuals have difficulty understanding log scale graphs, these issues do not tend to affect their support for intervention measures (perhaps in part because COVID-related news saturated the news and opinions were set outside of information provided in the experiments).

@@

If necessary or useful, information about certain computational details
such as version numbers, operating systems, or compilers could be included
in an unnumbered section. Also, auxiliary packages (say, for visualizations,
maps, tables, \dots) that are not cited in the main text can be credited here.
